---
title: "Analytics - Assignment 1"
subtitle: "**Cape Town Airbnb Listing Analysis**"
authors: "Ben Mitchell & Ben Frost"
format: html
editor: visual
---
```{r, echo = FALSE, warning = FALSE, message = FALSE}
# Required Package Installation:
library(tidyverse)
library(dplyr)
library(glmnetUtils)
library(randomForest)
library(gbm)
library(caret)
library(ggplot2)
```

```{r, echo = FALSE}
# Loading Data Set:
listings <- read.csv("listings.csv")
```

## Introduction:

The following report presents a supervised learning analysis of Airbnb listing data for the city of Cape Town. Based on any given listing observation the analysis aims to predict whether or not a host leases more than a threshold number of properties on Airbnb in which case they are termed a 'High Host' and explore this response's potential housing-market impact. Using a curated 2025 dataset of listings, this analysis develops and compares several classification models in order to identify patterns in listing characteristics that are associated with whether or not its host is considered a 'High Host'.

The analysis follows a structured predictive modelling workflow. The data are first divided into training and test sets so that model development and out of sample performance evaluation remain separate. The decision rule for what constitutes a 'High Host' is then derived and defined with the property number threshold being determined using the training data. Several classification models are subsequently fitted and tuned, the models types include elastic-net logistic regression, random forest, boosted trees and k-nearest neighbours. Their performance is compared using 10-fold cross-validation and a common set of evaluation metrics.

Beyond model comparison, this report also considers decision-threshold optimisation, since the best threshold number of properties for classification may differ from the default value when the objective is to improve the F1 score. The best performing model is selected and then interpreted using variable-importance and effect plots. Additionally, its prediction errors are examined through false-positive and false-negative cases. This report aims not only to assess predictive performance but also to demonstrate interpretable modelling processes.

```{r echo = FALSE}
# Training & Test Set Split (Raw Data)
set.seed(2026)

n <- nrow(listings)
trainIndex <- sample(seq_len(n), size = floor(0.8 * n))

trainDataC <- listings |> dplyr::slice(trainIndex)
testDataC  <- listings |> dplyr::slice(-trainIndex)
```

## Question 1:

This project will be modelling the binary target variable fitted to the Host Listings Count, the number of listings the host has (per Airbnb unknown calculations), above some variable to be defined, x.

There have been concerns in Cape Town regarding the massive influx of Airbnb short term rentals, which have had effects on the affordability and availability of long term-rentals. This binary target variable aims to discover what factors can describe listings that are attractive to professional Airbnb hosts who have portfolios of listings.

```{r echo = FALSE}
hostUnique <- trainDataC |> 
  distinct(host_id, host_listings_count)

HHtrain <- round(mean(hostUnique$host_listings_count > 2) * 100,2)
nHHtrain <- 100 - round(mean(hostUnique$host_listings_count > 2) * 100,2)
```

**Class Imbalance for x = 2 on Training Data:** <br> HighHost: `r HHtrain`% <br> Not HighHost: `r nHHtrain`% <br>

```{r echo = FALSE}
hostUnique <- testDataC |> 
  distinct(host_id, host_listings_count)

HHtest <- round(mean(hostUnique$host_listings_count > 2) * 100,2)
nHHtest <- 100 - round(mean(hostUnique$host_listings_count > 2) * 100,2)
```

**Class Imbalance for x = 2 on Test Data:** <br>HighHost: `r HHtest`% <br>Not HighHost: `r nHHtest`%

$$
\text{HighHost} =
\begin{cases}
1 & \text{if } \text{host_listings_count} > 2. \\
0 & \text{otherwise}
\end{cases}
$$

We have chosen to use 2 to be the threshold value used to define the binary target variable. This is appropriate as 27.4% of our training data shows hosts with more than 2 listings, and this is an indication of whether the host owns a portfolio of properties or if they are just listing their primary or secondary property.

```{r echo = FALSE}
hostCutoff = 2

# Binary Response: 1 = Yes, 0 = No
listings$host_listings_count <- ifelse(listings$host_listings_count > 2, 1, 0)

listings <- listings |>
  rename(HighHost = host_listings_count)
```

## Question 2:

```{r echo = FALSE}
# Training & Test Set Split With HighHost
set.seed(2026)

n <- nrow(listings)
trainIndex <- sample(seq_len(n), size = floor(0.8 * n))

trainData <- listings |> dplyr::slice(trainIndex)
testData  <- listings |> dplyr::slice(-trainIndex)
```

DATA PRE-PROCESSING:

```{r echo = FALSE}
# Remove id (Useless), host_id (Target Leakage Prevention) neighbourhood_ward (Collinear With Longitude & Latitude and Problematic):

  trainData <- trainData |>
    select(-id, -host_id, -neighbourhood_ward)
  
  testData <- testData |>
    select(-id, -host_id, -neighbourhood_ward)

# Convert host_since to a Usable Numeric Predictor:

  # Convert host_since to Date:
  trainData$host_since <- as.Date(trainData$host_since, format = "%Y/%m/%d")
  testData$host_since  <- as.Date(testData$host_since, format = "%Y/%m/%d")
  
  # Use Training Data to Define the Reference Date:
  minHostSince <- min(trainData$host_since)
  
  # Create Numeric Variable host_since_days:
  trainData$host_since_days <- as.numeric(trainData$host_since - minHostSince)
  testData$host_since_days  <- as.numeric(testData$host_since - minHostSince)
  
  # Remove host_since:
  trainData <- trainData |>
    select(-host_since)
  
  testData <- testData |>
    select(-host_since)
  
# Transform host_verifications into Usable Binary Features:
  
  # Binary Indicator Creation:
  trainData$has_email_verification <- ifelse(grepl("email", trainData$host_verifications), 1, 0)
  testData$has_email_verification  <- ifelse(grepl("email", testData$host_verifications), 1, 0)
  
  trainData$has_phone_verification <- ifelse(grepl("phone", trainData$host_verifications), 1, 0)
  testData$has_phone_verification  <- ifelse(grepl("phone", testData$host_verifications), 1, 0)
  
  trainData$has_work_email_verification <- ifelse(grepl("work_email", trainData$host_verifications), 1, 0)
  testData$has_work_email_verification  <- ifelse(grepl("work_email", testData$host_verifications), 1, 0)
  
  # Remove host_verifications:
  trainData <- trainData |>
    select(-host_verifications)
  
  testData <- testData |>
    select(-host_verifications)

# Transform Categorical Variables to be Factors:
  
  # *MIGHT NEED TO ENFORCE OBSERVATIONS CONTAINING NA DROPPING IN THE CASE THE MODEL ENCOUNTERS A FACTOR LEVEL IT HAS NEVER SEEN BEFORE WHEN BEING TESTED ON THE "UNSEEN" TEST DATA*
  
  # host_response_time:
  trainData$host_response_time <- as.factor(trainData$host_response_time)
  testData$host_response_time  <- factor(testData$host_response_time, levels = levels(trainData$host_response_time))
  
  # host_is_superhost
  trainData$host_is_superhost <- as.factor(trainData$host_is_superhost)
  testData$host_is_superhost  <- factor(testData$host_is_superhost, levels = levels(trainData$host_is_superhost))
  
  # room_type
  trainData$room_type <- as.factor(trainData$room_type)
  testData$room_type  <- factor(testData$room_type, levels = levels(trainData$room_type))
  
  # instant_bookable
  trainData$instant_bookable <- as.factor(trainData$instant_bookable)
  testData$instant_bookable  <- factor(testData$instant_bookable, levels = levels(trainData$instant_bookable))
  
# HighHost Variable Classes Required for Model Types:
  
  # Elastic Net : Numeric
  # Random Forest : Factor
  # GBM : Numeric
  # KNN : Factor
  
  # ENSURE DONE BEFORE MODEL CREATION
```

REGULARISED ELASTIC-NET LOGISTIC REGRESSION FITTING:

```{r, echo = FALSE}
elasticNetFile <- "elasticNetModel.RData"

#if (file.exists(elasticNetFile))
#{
#  load(elasticNetFile)
#} else 
#{
  set.seed(2026)
alphaGrid <- seq(0,1, by = 0.1)

elasticNetResults <- lapply(alphaGrid, function(a) {  
  elasticNetModel <- glmnetUtils::cv.glmnet(
    HighHost ~ .,
    data = trainData,
    family = "binomial",
    alpha = a,       # Tune Later
    nfolds = 10,
    type.measure = "auc"
  )
  
  index <- which.min(abs(elasticNetModel$lambda - elasticNetModel$lambda.min)) 
  
  data.frame(alpha = a, lambda_min = elasticNetModel$lambda.min, auc = elasticNetModel$cvm[index])
})
  elasticNetResults <- do.call(rbind, elasticNetResults)
  best_model_EN <- elasticNetResults[which.max(elasticNetResults$auc),]
  best_model_EN
  save(elasticNetModel, file = elasticNetFile) # Change to save the best model (may need to rerun the best model)
#}
```

REGULARISED ELASTIC-NET LOGISTIC REGRESSION INSPECTION:

```{r, echo = FALSE}
plot(x = elasticNetResults$alpha, y = elasticNetResults$auc, type = "l")
#print(elasticNetModel)
#coef(elasticNetModel, s = "lambda.min")
#plot(elasticNetModel)
```

CONVERTING HighHost TO BE A FACTOR:

```{r, echo = FALSE}
trainData$HighHost <- factor(trainData$HighHost, levels = c(0, 1), labels = c("No", "Yes"))
testData$HighHost  <- factor(testData$HighHost, levels = c(0, 1), labels = c("No", "Yes"))
```

RANDOM FOREST FITTING:

```{r, echo = FALSE}
#if (file.exists(randomForestFile)) {
#  load(randomForestFile)
#} else {

  set.seed(2026)

  # Number of predictors excluding response
  p <- ncol(trainData) - 1
  
  # Search space
  rfGrid <- expand.grid(
    mtry = c(floor(sqrt(p)), floor(p / 2)),
    ntree = seq(200, 500, 25),
    nodesize = c(2, 4, 7, 10)
  )
  
  # Fit all combinations and store final OOB error
  rfResults <- lapply(seq_len(nrow(rfGrid)), function(i) {
    
    rfFit <- randomForest(
      HighHost ~ .,
      data = trainData,
      ntree = rfGrid$ntree[i],
      mtry = rfGrid$mtry[i],
      nodesize = rfGrid$nodesize[i],
      importance = TRUE
    )
    
    data.frame(
      mtry = rfGrid$mtry[i],
      ntree = rfGrid$ntree[i],
      nodesize = rfGrid$nodesize[i],
      OOB_Error = rfFit$err.rate[nrow(rfFit$err.rate), "OOB"]
    )
  })
#  save(randomForestModel, file = randomForestFile)
#}
```

RANDOM FOREST FITTING INSPECTION:

```{r, echo = FALSE}
rfResults <- do.call(rbind, rfResults)

# Find best combination
best_model_rf <- rfResults[which.min(rfResults$OOB_Error), ]
best_model_rf

# Plot tuning results
ggplot(rfResults, aes(x = ntree, y = OOB_Error,
                      colour = factor(nodesize),
                      group = factor(nodesize))) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ mtry, scales = "fixed") +
  labs(
    title = "Random forest tuning results",
    x = "Number of trees",
    y = "OOB Error",
    colour = "nodesize"
  ) +
  theme_minimal()
#randomForestModel
#plot(randomForestModel$err.rate[, "OOB"], type = "l", xlab = "Number of Trees", ylab = "OOB Error", col = "blue", lwd = 2, ylim = c(0, max(randomForestModel$err.rate[, "OOB"])), main = "Random Forest Model: OOB Error vs Number of Trees")
#print(randomForestModel)
#importance(randomForestModel)
#varImpPlot(randomForestModel)
```
### GBM Boosted Tree Model:

For the GBM boosted tree model, a gradient boosting classifier with a Bernoulli loss function was fitted to the training data, since the response variable HighHost is binary. The model builds many small decision trees sequentially, with each new tree focusing on correcting the errors made by the previous ensemble. This allows the method to capture non-linear effects and interactions among the predictors while still being tuned in a controlled way.

Hyperparameter selection was carried out using a grid-search approach on the training set only. For each candidate combination of tuning parameters, a GBM model was fitted with 10-fold cross-validation and model performance was assessed using the cross-validated Bernoulli deviance. The preferred hyperparameter combination was the one that achieved the lowest cross-validated error without risk of overfitting to the training data. After identifying the best combination of values for these hyperparameters the final GBM model has been refitted using these values.

The hyperparameter search space was:

n.trees ∈ {500, 1000, 1500, 2000}

interaction.depth ∈ {1, 3, 5}

shrinkage ∈ {0.005, 0.01, 0.05}

n.minobsinnode ∈ {5, 10, 20}

This gives a total of 4 × 3 × 3 × 3 = 108 hyperparameter combinations evaluated.


```{r}
# CONVERTING HighHost TO BE NUMERIC:
trainData$HighHost <- ifelse(trainData$HighHost == "Yes", 1, 0)
testData$HighHost  <- ifelse(testData$HighHost == "Yes", 1, 0)
```

GBM BOOSTED TREE MODEL FITTING:

```{r}
#if (file.exists(gbmFile)) {
#  load(gbmFile)
#} else {
  set.seed(2026)

  gbmGrid <- expand.grid(
  n.trees = c(500, 1000, 1500, 2000),
  interaction.depth = c(1, 3, 5),
  shrinkage = c(0.005, 0.01, 0.05),
  n.minobsinnode = c(5, 10, 20))
  
  gbmResults <- lapply(seq_len(nrow(gbmGrid)), function(i) {
  
  gbmFit <- gbm(
    formula = HighHost ~ .,
    data = trainData,
    distribution = "bernoulli",
    n.trees = gbmGrid$n.trees[i],
    interaction.depth = gbmGrid$interaction.depth[i],
    shrinkage = gbmGrid$shrinkage[i],
    n.minobsinnode = gbmGrid$n.minobsinnode[i],
    cv.folds = 10,
    verbose = FALSE
  )
  
  bestIter <- gbm.perf(gbmFit, method = "cv", plot.it = FALSE)
  bestCvError <- gbmFit$cv.error[bestIter]
  
  data.frame(
    n.trees = gbmGrid$n.trees[i],
    interaction.depth = gbmGrid$interaction.depth[i],
    shrinkage = gbmGrid$shrinkage[i],
    n.minobsinnode = gbmGrid$n.minobsinnode[i],
    best_iter = bestIter,
    CV_Error = bestCvError
  )
})

gbmResults <- do.call(rbind, gbmResults)

best_model_gbm <- gbmResults[which.min(gbmResults$CV_Error), ]
best_model_gbm

save(gbmModel, gbmResults, best_model_gbm, bestTrees, file = "gbm_all_objects.RData")
```

GBM BOOSTED TREE MODEL TUNING:

```{r}
ggplot(
  gbmResults,
  aes(
    x = n.trees,
    y = CV_Error,
    colour = factor(n.minobsinnode),
    group = factor(n.minobsinnode)
  )
) +
  geom_line() +
  geom_point(size = 2) +
  facet_grid(shrinkage ~ interaction.depth, labeller = label_both) +
  labs(
    title = "GBM Tuning Results",
    x = "Number of trees",
    y = "Cross-Validated Bernoulli Deviance",
    colour = "minobsinnode"
  ) +
  theme_minimal()

```

### K-Nearest Neighbours Model:

For the K-nearest neighbours model, a supervised classification approach was used in which each observation is classified according to the majority class among its k nearest neighbours in the predictor space. Because KNN is distance-based, predictor scaling is important, so all predictors were centred and scaled before model fitting. Model selection was performed on the training set only using 10-fold cross-validation, with the optimal value of k chosen as the one that maximised the cross-validated ROC measure.

In addition to tuning the number of neighbours, the model was also tuned over alternative candidate feature subsets. This was done because KNN can be highly sensitive to irrelevant or weak predictors: including too many variables can distort the distance calculation and reduce classification performance. A range of feature sets was therefore considered, covering core listing characteristics, host-related variables, review-related variables, availability measures, location variables, and broader combined sets. The final KNN specification was chosen as the combination of feature subset and k with the highest mean cross-validated ROC.

The feature search space consisted of the following candidate subsets:

Basic: price, accommodates, bathrooms, bedrooms, beds, minimum_nights, maximum_nights

Host: host_since_days, host_response_rate, host_acceptance_rate, host_response_time, host_is_superhost, has_email_verification, has_phone_verification, has_work_email_verification

Review: number_of_reviews, reviews_per_month, review_scores_rating

Availability: availability_30, availability_60, availability_90, availability_365

Location: latitude, longitude, room_type, instant_bookable

Combined 1:
price, accommodates, bathrooms, bedrooms, beds, minimum_nights, maximum_nights,
host_since_days, host_response_rate, host_acceptance_rate, host_is_superhost,
number_of_reviews, reviews_per_month, review_scores_rating,
latitude, longitude, room_type, instant_bookable

Combined 2:
price, accommodates, bathrooms, bedrooms, beds, minimum_nights, maximum_nights,
host_since_days, host_response_rate, host_acceptance_rate, host_response_time, host_is_superhost,
has_email_verification, has_phone_verification, has_work_email_verification,
number_of_reviews, reviews_per_month, review_scores_rating,
availability_30, availability_60, availability_90, availability_365,
latitude, longitude, room_type, instant_bookable

The neighbourhood size search space was:

k∈ {3,5,7,…,25}

So the KNN tuning process searched jointly across 7 candidate feature sets and 12 candidate values of k.

```{r, echo = FALSE}
# CONVERTING HighHost TO BE A FACTOR:
trainData$HighHost <- factor(trainData$HighHost, levels = c(0, 1), labels = c("No", "Yes"))
testData$HighHost  <- factor(testData$HighHost, levels = c(0, 1), labels = c("No", "Yes"))
```

K-NEAREST NEIGHBOURS FITTING:

```{r}
# REMOVE FEATURES WITH ZERO VARIANCE

featureSets <- list(
  Basic = c(
    "price", "accommodates", "bathrooms", "bedrooms", "beds",
    "minimum_nights", "maximum_nights"
  ),
  Host = c(
    "host_since_days", "host_response_rate", "host_acceptance_rate",
    "host_response_time", "host_is_superhost",
    "has_email_verification", "has_phone_verification",
    "has_work_email_verification"
  ),
  Review = c(
    "number_of_reviews", "reviews_per_month", "review_scores_rating"
  ),
  Availability = c(
    "availability_30", "availability_60", "availability_90", "availability_365"
  ),
  Location = c(
    "latitude", "longitude", "room_type", "instant_bookable"
  ),
  Combined_1 = c(
    "price", "accommodates", "bathrooms", "bedrooms", "beds",
    "minimum_nights", "maximum_nights",
    "host_since_days", "host_response_rate", "host_acceptance_rate",
    "host_is_superhost",
    "number_of_reviews", "reviews_per_month", "review_scores_rating",
    "latitude", "longitude", "room_type", "instant_bookable"
  ),
  Combined_2 = c(
    "price", "accommodates", "bathrooms", "bedrooms", "beds",
    "minimum_nights", "maximum_nights",
    "host_since_days", "host_response_rate", "host_acceptance_rate",
    "host_response_time", "host_is_superhost",
    "has_email_verification", "has_phone_verification",
    "has_work_email_verification",
    "number_of_reviews", "reviews_per_month", "review_scores_rating",
    "availability_30", "availability_60", "availability_90", "availability_365",
    "latitude", "longitude", "room_type", "instant_bookable"
  )
)

#knnFile <- "knnModel.RData"

#if (file.exists(knnFile)) {
#  load(knnFile)
#} else {
  set.seed(2026)

  knnControl <- trainControl(
    method = "cv",
    number = 10,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    verboseIter = TRUE
  )
  
  knnGrid <- expand.grid(k = seq(1, 25, by = 2))

  knnFits <- list()
  knnResults <- list()

  for (setName in names(featureSets)) 
  {
  
    vars <- featureSets[[setName]]
    
    trainSubset <- trainData[ , c(vars, "HighHost")]
    
    knnFit <- train(
      HighHost ~ .,
      data = trainSubset,
      method = "knn",
      trControl = knnControl,
      tuneGrid = knnGrid,
      metric = "ROC",
      preProcess = c("center", "scale")
    )
    
    knnFits[[setName]] <- knnFit
    
    temp <- knnFit$results
    temp$FeatureSet <- setName
    knnResults[[setName]] <- temp
  }

knnResults <- do.call(rbind, knnResults)

best_knn <- knnResults[which.max(knnResults$ROC), ]
best_knn

save(
  knnFits,
  knnResults,
  best_knn,
  featureSets,
  knnGrid,
  file = "knn_all_objects.RData"
)

#  save(knnModel, file = knnFile)
#}
```

K-NEAREST NEIGHBOURS INSPECTION:

```{r}
load("knn_all_objects.RData")

ggplot(
  knnResults,
  aes(
    x = k,
    y = ROC,
    colour = FeatureSet,
    group = FeatureSet
  )
) +
  geom_line() +
  geom_point(size = 2) +
  labs(
    title = "K-Nearest Neighbours Tuning Results",
    x = "Number of Neighbours (k)",
    y = "Mean CV ROC",
    colour = "Feature Set"
  ) +
  theme_minimal()

print(best_knn)
plot(best_knn)
```


